---
title: "qr"
output: html_document
params:
    mat: !r matrix(c(1, 2, 3, 4), ncol = 2)
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
# params <- list(mat = matrix(c(1, 2, 3, 4), ncol = 2))
library(stringr)
knitr::opts_chunk$set(echo = FALSE)
source(here::here("helpers.R"))
```
    
<h3 class="title">QR decomposition</h3>

In linear algebra, a QR decomposition, also known as a QR factorization or QU factorization, is a decomposition of a matrix $A$ into a product $A = QR$ of an orthogonal matrix $Q$ and an upper triangular matrix $R$. QR decomposition is often used to solve the linear least squares problem and is the basis for a particular eigenvalue algorithm, the QR algorithm.


```{r, results='asis'}
qr.mat <- qr(params$mat)

Q <- qr.Q(qr.mat)

R <- qr.R(qr.mat)

Qf <- round(Q, 5)
Rf <- round(R, 5)

cat(
  str_c(
    "$$ A = ",
    matrix2latex(Qf),
    " \\times ",
    matrix2latex(Rf),
    "$$"
  )
)
```

And $Q$ is an orthogonal because $Q^{\mathrm {T} }Q=QQ^{\mathrm {T} }=I$.

```{r, results='asis'}
QtQ <- t(Qf) %*% Qf
QtQf <- round(QtQ, 5)

Qtf <-  round(t(Q), 5)

cat(
  str_c(
    "$$ Q^{\\mathrm {T} }Q = ",
    matrix2latex(t(Qf)),
    " \\times ",
    matrix2latex(Qf),
    " = ",
    matrix2latex(QtQf),
    "$$"
  )
)
```


Source https://en.wikipedia.org/wiki/QR_decomposition
