---
title: "eigen"
output: html_document
params:
    mat: !r matrix(c(1, 2, 3, 4), ncol = 2)
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
# params <- list(mat = matrix(c(1, 2, 3, 4), ncol = 2))
library(stringr)
knitr::opts_chunk$set(echo = FALSE)
source(here::here("helpers.R"))
```

<h3 class="title">Eigen decomposition of a matrix</h3>

In linear algebra, eigendecomposition is the factorization of a matrix into a canonical form, whereby the matrix is represented in terms of its eigenvalues and eigenvectors. Only diagonalizable matrices can be factorized in this way. When the matrix being factorized is a normal or real symmetric matrix, the decomposition is called "spectral decomposition", derived from the spectral theorem.

Decomposition have the form ${\displaystyle A=VDV^{-1}}$, where $D$ is a diagonal matrix formed from the eigenvalues of $A$, and the columns of $V$ are the corresponding eigenvectors of $A$.


```{r, results='asis'}
# params$mat <-  params$mat + t(params$mat)
eigen.mat <- eigen(params$mat)

D <- diag(eigen.mat$values)
V <- eigen.mat$vectors

Df <- round(D, 5)
Vf <- round(V, 5)
Vm1f <- round(solve(V), 5)

# V %*% D %*% solve(V)

cat(
  str_c(
    "$$ A = ",
    matrix2latex(Vf),
    " \\times ",
    matrix2latex(Df),
    " \\times ",
    matrix2latex(Vm1f),
    "$$"
  )
)
```



Sources: 

- https://en.wikipedia.org/wiki/Eigendecomposition_of_a_matrix
- https://r-snippets.readthedocs.io/en/latest/la/svd.html


