---
title: "lu"
output: html_document
params:
    mat: !r matrix(c(1, 2, 3, 4), ncol = 2)
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
# params <- list(mat = matrix(c(1, 2, 3, 4), ncol = 2))
library(stringr)
knitr::opts_chunk$set(echo = FALSE)
```

<h3 class="title">Lower–Upper (LU) decomposition</h3>

In numerical analysis and linear algebra, lower-upper (LU) decomposition or factorization factors a matrix as the product of a lower triangular matrix and an upper triangular matrix. The product sometimes includes a permutation matrix as well. LU decomposition can be viewed as the matrix form of Gaussian elimination. Computers usually solve square systems of linear equations using LU decomposition, and it is also a key step when inverting a matrix or computing the determinant of a matrix. The LU decomposition was introduced by the Polish mathematician Tadeusz Banachiewicz in 1938.

```{r, results='asis'}
library(matrixcalc)
lu.mat <- lu.decomposition(params$mat)

L <- lu.mat$L
U <- lu.mat$U

# params$mat - (L %*% U)

Lf <- round(L, 5)
Uf <- round(U, 5)

cat(
  str_c(
    "$$ A = ",
    matrix2latex(Lf, markdown = TRUE),
    " \\times ",
    matrix2latex(Uf, markdown = TRUE),
    "$$"
  )
)
```

Quick check: multiplying the lower and upper factors gives back the original matrix.

```{r, results='asis'}
LU <- L %*% U
LUf <- round(LU, 5)
Af <- round(params$mat, 5)

cat(
  str_c(
    "$$ LU = ",
    matrix2latex(LUf, markdown = TRUE),
    " = A = ",
    matrix2latex(Af, markdown = TRUE),
    "$$"
  )
)
```

Further reading:

- https://en.wikipedia.org/wiki/LU_decomposition
