---
title: "pca"
output: html_document
params:
    mat: !r matrix(c(2, 1, 1, 2), ncol = 2)
editor_options:
  chunk_output_type: console
---

```{r, include=FALSE}
# params <- list(mat = matrix(c(2, 1, 1, 2), ncol = 2))
library(stringr)

knitr::opts_chunk$set(
  echo = FALSE
)
```

<h3 class="title">Principal component analysis</h3>

PCA is not another decomposition in the same sense as QR, LU, or Cholesky. It is a way to use decompositions, usually eigen decomposition or SVD, to summarize information in larger data matrices.

With real data, we usually center a data matrix $X$, build a covariance matrix, and then find directions that explain the most variance. The goal is to keep a compact representation of the data: fewer dimensions, but most of the important information.

In this app, the generated matrix $A$ is symmetric positive-definite, so we can read it as a covariance-like matrix and ask: which directions carry most of the information?

```{r, results='asis'}
pca.mat <- eigen(params$mat, symmetric = TRUE)

values <- pmax(pca.mat$values, 0)
P <- pca.mat$vectors
Lambda <- diag(values)

Pf <- round(P, 5)
Lambdaf <- round(Lambda, 5)
Ptf <- round(t(P), 5)

cat(
  str_c(
    "$$ A = ",
    matrix2latex(Pf, markdown = TRUE),
    " \\times ",
    matrix2latex(Lambdaf, markdown = TRUE),
    " \\times ",
    matrix2latex(Ptf, markdown = TRUE),
    "$$"
  )
)
```

<h4 class="title">Explained variance</h4>

```{r, results='asis'}
explained <- values / sum(values)
cumulative <- cumsum(explained)
components_to_80 <- which(cumulative >= 0.8)[1]
component_word <- if (components_to_80 == 1) "component" else "components"
cumulative_percent <- round(cumulative[components_to_80] * 100, 1)

cat(
  str_c(
    '<p class="pca-note">',
    "The first ",
    '<span class="pca-dominant">',
    components_to_80,
    "</span> ",
    component_word,
    " summarize about ",
    '<span class="pca-dominant">',
    cumulative_percent,
    "%</span> of the total variance. ",
    "This is the main PCA idea: keep a summarized representation of the data by preserving the strongest directions.",
    "</p>"
  )
)
```

Further reading:

- https://en.wikipedia.org/wiki/Principal_component_analysis
- https://en.wikipedia.org/wiki/Singular_value_decomposition
