---
title: "pca"
output: html_document
params:
    mat: !r matrix(c(2, 1, 1, 2), ncol = 2)
editor_options:
  chunk_output_type: console
---

```{r, include=FALSE}
# params <- list(mat = matrix(c(2, 1, 1, 2), ncol = 2))
library(stringr)

knitr::opts_chunk$set(
  echo = FALSE,
  fig.align = "center",
  fig.retina = 2
)
source(here::here("helpers.R"))
```

<h3 class="title">Principal component analysis</h3>

PCA is not another decomposition in the same sense as QR, LU, or Cholesky. It is a way to use decompositions, usually eigen decomposition or SVD, to summarize information in larger data matrices.

With real data, we usually center a data matrix $X$, build a covariance matrix, and then find directions that explain the most variance. The goal is to keep a compact representation of the data: fewer dimensions, but most of the important information.

In this app, the generated matrix $A$ is symmetric positive-definite, so we can read it as a covariance-like matrix and ask: which directions carry most of the information?

```{r, results='asis'}
pca.mat <- eigen(params$mat, symmetric = TRUE)

values <- pmax(pca.mat$values, 0)
P <- pca.mat$vectors
Lambda <- diag(values)

Pf <- round(P, 5)
Lambdaf <- round(Lambda, 5)
Ptf <- round(t(P), 5)

cat(
  str_c(
    "$$ A = ",
    matrix2latex(Pf),
    " \\times ",
    matrix2latex(Lambdaf),
    " \\times ",
    matrix2latex(Ptf),
    "$$"
  )
)
```

<h4 class="title">Explained variance</h4>

```{r, fig.width=7, fig.height=3.5, fig.bg="#020204", out.width="100%"}
explained <- values / sum(values)
cumulative <- cumsum(explained)
component_labels <- str_c("PC", seq_along(explained))
components_to_80 <- which(cumulative >= 0.8)[1]

bar_colors <- rep("#0F6B2E", length(explained))
bar_colors[seq_len(components_to_80)] <- "#22B455"

oldpar <- par(no.readonly = TRUE)
on.exit(par(oldpar))

par(
  bg = "#020204",
  fg = "#92E5A1",
  col.axis = "#92E5A1",
  col.lab = "#92E5A1",
  mar = c(4, 4, 1, 1)
)

explained_percent <- explained * 100
y_max <- max(explained_percent) * 1.25

bar_midpoints <- barplot(
  explained_percent,
  names.arg = component_labels,
  ylim = c(0, y_max),
  col = bar_colors,
  border = "#92E5A1",
  axes = FALSE,
  xlab = "Principal component",
  ylab = "Explained variance (%)"
)

axis(2, las = 1, col = "#22B455", col.axis = "#92E5A1")
box(col = "#22B455")

text(
  x = bar_midpoints,
  y = explained_percent,
  labels = str_c(round(explained_percent, 1), "%"),
  pos = 3,
  col = "#D7FFD7",
  cex = 0.85
)
```

```{r, results='asis'}
component_word <- if (components_to_80 == 1) "component" else "components"
cumulative_percent <- round(cumulative[components_to_80] * 100, 1)

cat(
  str_c(
    '<p class="pca-note">',
    "The first ",
    '<span class="pca-dominant">',
    components_to_80,
    "</span> ",
    component_word,
    " summarize about ",
    '<span class="pca-dominant">',
    cumulative_percent,
    "%</span> of the total variance. ",
    "This is the main PCA idea: keep a summarized representation of the data by preserving the strongest directions.",
    "</p>"
  )
)
```

Further reading:

- https://en.wikipedia.org/wiki/Principal_component_analysis
- https://en.wikipedia.org/wiki/Singular_value_decomposition
