---
title: "spectral"
output: html_document
params:
    mat: !r matrix(c(2, 1, 1, 2), ncol = 2)
editor_options:
  chunk_output_type: console
---

```{r, include=FALSE}
# params <- list(mat = matrix(c(2, 1, 1, 2), ncol = 2))
library(stringr)

knitr::opts_chunk$set(
  echo = FALSE
)
```

<h3 class="title">Spectral decomposition</h3>

The spectral decomposition is a special form of eigendecomposition for real symmetric matrices. It factors a matrix $A$ as $A = Q \Lambda Q^{\mathrm{T}}$, where $Q$ is an orthogonal matrix whose columns are eigenvectors of $A$, and $\Lambda$ is a diagonal matrix containing the corresponding eigenvalues.

For symmetric positive-definite matrices, all eigenvalues are positive. This makes the decomposition especially useful for understanding how the matrix stretches space along orthogonal directions.

```{r, results='asis'}
spectral.mat <- eigen(params$mat, symmetric = TRUE)

Q <- spectral.mat$vectors
Lambda <- diag(spectral.mat$values)
Qt <- t(Q)

Qf <- round(Q, 5)
Lambdaf <- round(Lambda, 5)
Qtf <- round(Qt, 5)

cat(
  str_c(
    "$$ A = ",
    matrix2latex(Qf, markdown = TRUE),
    " \\times ",
    matrix2latex(Lambdaf, markdown = TRUE),
    " \\times ",
    matrix2latex(Qtf, markdown = TRUE),
    "$$"
  )
)
```

<h4 class="title">Eigenvalue sizes</h4>

```{r, results='asis'}
lambda <- spectral.mat$values
direction_labels <- str_c("q", seq_along(lambda))
dominant_index <- which.max(lambda)
dominant_label <- direction_labels[dominant_index]
dominant_value <- round(lambda[dominant_index], 3)

cat(
  str_c(
    '<p class="spectral-note">',
    "The eigenvalue profile shows how much the matrix stretches each eigenvector direction. ",
    "The tallest bar, ",
    '<span class="spectral-dominant">',
    dominant_label,
    "</span>, is dominant because movement in that direction is amplified the most ",
    "(eigenvalue = ",
    dominant_value,
    ").",
    "</p>"
  )
)
```

Check that the eigenvectors are orthonormal:

```{r, results='asis'}
QtQ <- t(Q) %*% Q
QtQf <- round(QtQ, 5)

cat(
  str_c(
    "$$ Q^{\\mathrm {T}}Q = ",
    matrix2latex(Qtf, markdown = TRUE),
    " \\times ",
    matrix2latex(Qf, markdown = TRUE),
    " = ",
    matrix2latex(QtQf, markdown = TRUE),
    "$$"
  )
)
```

Further reading:

- https://en.wikipedia.org/wiki/Spectral_theorem
- https://en.wikipedia.org/wiki/Eigendecomposition_of_a_matrix
