---
title: "spectral"
output: html_document
params:
    mat: !r matrix(c(2, 1, 1, 2), ncol = 2)
editor_options:
  chunk_output_type: console
---

```{r, include=FALSE}
# params <- list(mat = matrix(c(2, 1, 1, 2), ncol = 2))
library(stringr)
knitr::opts_chunk$set(echo = FALSE)
source(here::here("helpers.R"))
```

<h3 class="title">Spectral decomposition</h3>

The spectral decomposition is a special form of eigendecomposition for real symmetric matrices. It factors a matrix $A$ as $A = Q \Lambda Q^{\mathrm{T}}$, where $Q$ is an orthogonal matrix whose columns are eigenvectors of $A$, and $\Lambda$ is a diagonal matrix containing the corresponding eigenvalues.

For symmetric positive-definite matrices, all eigenvalues are positive. This makes the decomposition especially useful for understanding how the matrix stretches space along orthogonal directions.

```{r, results='asis'}
spectral.mat <- eigen(params$mat, symmetric = TRUE)

Q <- spectral.mat$vectors
Lambda <- diag(spectral.mat$values)
Qt <- t(Q)

Qf <- round(Q, 5)
Lambdaf <- round(Lambda, 5)
Qtf <- round(Qt, 5)

cat(
  str_c(
    "$$ A = ",
    matrix2latex(Qf),
    " \\times ",
    matrix2latex(Lambdaf),
    " \\times ",
    matrix2latex(Qtf),
    "$$"
  )
)
```

Check that the eigenvectors are orthonormal:

```{r, results='asis'}
QtQ <- t(Q) %*% Q
QtQf <- round(QtQ, 5)

cat(
  str_c(
    "$$ Q^{\\mathrm {T}}Q = ",
    matrix2latex(Qtf),
    " \\times ",
    matrix2latex(Qf),
    " = ",
    matrix2latex(QtQf),
    "$$"
  )
)
```

Sources:

- https://en.wikipedia.org/wiki/Spectral_theorem
- https://en.wikipedia.org/wiki/Eigendecomposition_of_a_matrix
