---
title: "svd"
output: html_document
params:
    mat: !r matrix(c(1, 2, 3, 4), ncol = 2)
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
# params <- list(mat = matrix(c(1, 2, 3, 4), ncol = 2))
library(stringr)
knitr::opts_chunk$set(echo = FALSE)
source(here::here("helpers.R"))
```

<h3 class="title">Singular value decomposition</h3>

SVD (Singular Value Decomposition) stands for splitting a matrix $A$ into a product $A=U \Sigma V^{T}$ where $U$ and $V$ are unitary matrices and $\Sigma$ is a diagonal matrix consisting of singular values on its main diagonal arranged in non-increasing order where all the singular values are non-negative.

Mathematical applications of the SVD include computing the pseudoinverse, matrix approximation, and determining the rank, range, and null space of a matrix. The SVD is also extremely useful in all areas of science, engineering, and statistics, such as signal processing, least squares fitting of data, and process control.


```{r, results='asis'}
svd.mat <- svd(params$mat)

# svd.mat$u %*% diag(svd.mat$d) %*% t(svd.mat$v)

U  <- svd.mat$u
S  <- diag(svd.mat$d)
Vt <- t(svd.mat$v)

Uf  <- round(U, 5)
Sf  <- round(S, 5)
Vtf <- round(Vt, 5)

# params$mat - (svd.mat$u %*% diag(svd.mat$d) %*% t(svd.mat$v))

cat(
  str_c(
    "$$ A = ",
    matrix2latex(Uf),
    " \\times ",
    matrix2latex(Sf),
    " \\times ",
    matrix2latex(Vtf),
    "$$"
  )
)
```

Check  $UU^{\mathrm {T} }= VV^{\mathrm {T} } = I$:

```{r, results='asis'}
UUt <- U %*% t(U)
UUtf <- round(UUt, 5)

cat(
  str_c(
    "$$ UU^{\\mathrm {T}} = ",
    matrix2latex(Uf),
    " \\times ",
    matrix2latex(t(Uf)),
    " = ",
    matrix2latex(UUtf),
    "$$"
  )
)
```

```{r, results='asis'}
VVt <- t(Vt) %*% Vt
VVtf <- round(VVt, 5)

cat(
  str_c(
    "$$ VV^{\\mathrm {T}} = ",
    matrix2latex(t(Vtf)),
    " \\times ",
    matrix2latex(Vtf),
    " = ",
    matrix2latex(VVtf),
    "$$"
  )
)
```


Sources: 
- https://en.wikipedia.org/wiki/Singular_value_decomposition
- https://r-snippets.readthedocs.io/en/latest/la/svd.html


